On the power of oritatami cotranscriptional folding with unary bead sequence\thanks{This work is supported in part by KAKENHI Grant-in-Aid for Challenging Research (Exploratory) No.~18K19779 and JST Program to Disseminate Tenure Tracking System No.~6F36 granted to S.~Z.~F. and S.~S.}

### 2018-12-06 arity1upper.tex
Technical problems fixed

\subsection{$\alpha = 1$, unary}
\label{sec:unary}

In this section we will argue that unary systems at arity $1$ and delay $2$ and $3$, respectively, cannot fold infinite transcripts deterministically. As we will see, in fact, the length of transcripts deterministically foldable by these systems has an upper bound quadratic in the length of the seed. The main result is the following theorem, which is a direct consequence of Lemma~\ref{lem:d2a1}  and Lemma~\ref{lem:d3a1} which follow.
\begin{theorem}[$\delta\in\{ 2,3\}, \alpha = 1$]\label{thm:d23a1}
	Let $\Xi$ be a unary oritatami system of $\delta \in \{2,3\}$ and $\alpha = 1$. Starting with a seed of length $n$, $\Xi$ can yield only finite structures of size at most $3n^2+4n+2$.
\end{theorem}

Let us fix some common starting points for the next two subsections. Let the point where the first transcript bead was fixed be $p$ and let $n=|\mathrm{seed}|+1$. We will argue about the situation when the first bead is stabilized outside $\hexagon_p^n$ (a hexagon of radius $n$). Let this be the $i$th bead of the transcript. Without loss of generality, we can translate the origin $(0,0)$ to the coordinates of bead $i-1$ (which is still in $\hexagon_p^n$), and we can assume that bead $i$ is fixed at $(1,1)$ (see Fig.~\ref{fig:hexagonOut}).
\begin{figure}
	\centering
	\includegraphics[width=0.3\linewidth]{./Fig/hexagonOut}
	\caption{$\hexagon_p^n$ and the position $(1,1)$ of the first bead fixed outside of it.}
	\label{fig:hexagonOut}
\end{figure}

%\subsubsection{\ref{sec:unary}.1 \ \ $\delta = 2$\\}
\begin{lemma}[$\delta=2, \alpha =1$]\label{lem:d2a1}
	Let $\Xi$ be a unary oritatami system of $\delta=2$ and $\alpha = 1$. Starting with a seed of length $n$, $\Xi$ can yield only finite structures of size at most $3n^2+4n+2$.
\end{lemma}
\begin{proof}

In the elongation that places bead $i$ at $(1,1)$ there are two possibilities:
\begin{itemize}
	\item $i$ forms a bond with a bead at $(1,0)$.
	\item  $i$ does not bond to anything and $i+1$ is at $(2,1)$ bonding with a bead at $(2,0)$. If there is no bead at $(1,0)$, then placing $i$ at $(1,0)$ instead of $(1,1)$ results in the same number of bonds, leading to nondeterminism. Therefore, there has to be a bead at $(1,0)$ and it is inactive, otherwise it would bond to $i$. This is analogous to case 1. below, with the only difference being that the bond between (1,1) and (1,0) is missing.%as in Fig.~\ref{fig:hexagonOut1}.
	\begin{figure}
		\centering
		\includegraphics[width=0.2\linewidth]{./Fig/hexagonOut1}
		%\caption{}
		\label{fig:hexagonOut1}
	\end{figure}

\end{itemize}

@@

\begin{figure}
	\centering
	%\includegraphics[width=0.3\linewidth]{./hexagonOut1}
	%\hspace{10mm} %
	\includegraphics[width=0.8\linewidth]{./Fig/hexagonOut3n}

@@

Case 2.(a) gives the same upper bound as case 1. Cases 2.(b)-(c) give the smaller upper bound $|\hexagon_p^n|+1$, thereby concluding the proof.
\end{proof}

@@

\begin{lemma}[$\delta=3, \alpha =1$]\label{lem:d3a1}
	Let $\Xi$ be a unary oritatami system of $\delta=3$ and $\alpha = 1$. Starting with a seed of length $n$, $\Xi$ can yield only finite structures of size at most $3n^2+4n+2$.
\end{lemma}
\begin{proof}
We will argue similarly to the $\delta=2$ case: when we stabilize beads outside $\hexagon_p^n$, we can only do so at points right next to one of the sides and even there only until we reach a corner. This yields the upper bound immediately.

>>>>

\subsection{Quadratic upper bounds for arity 1 and delay 2 or 3}
\label{sec:unary}

In this section we will argue that unary systems at arity $1$ and delay $2$ and $3$, respectively, cannot fold infinite transcripts deterministically. As we will see, in fact, the length of transcripts deterministically foldable by these systems has an upper bound quadratic in the length $n$ of the seed. The main result is the following theorem, which is a direct consequence of Lemmas~\ref{lem:d2a1}  and \ref{lem:d3a1} which follow.
\begin{theorem}[$\delta\in\{ 2,3\}, \alpha = 1$]\label{thm:d23a1}
	A deterministic unary oritatami system at arity $\alpha = 1$ and delay $\delta \in \{2,3\}$ can fold into a terminal conformation of length at most $3n^2+4n+2$.
\end{theorem}

%\begin{figure}
\begin{wrapfigure}{r}{0.5\linewidth}
	\centering
	\includegraphics[width=0.4\linewidth]{./Fig/hexagonOut}
	\caption{$\hexagon_p^n$ and the position $(1,1)$ of the first bead fixed outside of it.}
	\label{fig:hexagonOut}
\end{wrapfigure}
%\end{figure}

Let us fix some common starting points for Lemmas~\ref{lem:d2a1} and \ref{lem:d3a1}. Let the point where the first transcript bead was fixed be $p$ and let $n=|\mathrm{seed}|+1$. We will argue about the situation when the first bead is stabilized outside $\hexagon_p^n$ (a hexagon of radius $n$). Let this be the $i$th bead of the transcript. Without loss of generality, we can translate the origin $(0,0)$ to the coordinates of bead $i-1$ (which is still in $\hexagon_p^n$), and we can assume that bead $i$ is fixed at $(1,1)$ (see Fig.~\ref{fig:hexagonOut}).

%\subsubsection{\ref{sec:unary}.1 \ \ $\delta = 2$\\}
\begin{lemma}[$\delta=2, \alpha =1$]\label{lem:d2a1}
	A deterministic unary oritatami system of $\delta=2$ and $\alpha = 1$ can fold into a conformation of length at most $3n^2+4n+2$.
\end{lemma}
\begin{proof}
In the elongation that places bead $i$ at $(1,1)$ there are two possibilities. 
\begin{itemize}
	\item $i$ forms a bond with a bead at $(1,0)$.
	\item  $i$ does not bond to anything and $i+1$ is at $(2,1)$ bonding with a bead at $(2,0)$. If there is no bead at $(1,0)$, then placing $i$ at $(1,0)$ instead of $(1,1)$ results in the same number of bonds, leading to nondeterminism. Therefore, there has to be a bead at $(1,0)$ and it is inactive, otherwise it would bond to $i$. This is analogous to case 1. below, with the only difference being that the bond between (1,1) and (1,0) is missing.%as in Fig.~\ref{fig:hexagonOut1}.
\end{itemize}

	\begin{figure}
		\centering
		\includegraphics[width=0.2\linewidth]{./Fig/hexagonOut1}
		%\caption{}
		\label{fig:hexagonOut1}
	\end{figure}

@@

\begin{figure}[tb]
	\centering
	%\includegraphics[width=0.3\linewidth]{./hexagonOut1}
	%\hspace{10mm} %
	\includegraphics[width=0.8\linewidth]{./Fig/hexagonOut3n}

@@

Case 2.(a) gives the same upper bound as case 1. Cases 2.(b)-(c) give the smaller upper bound $|\hexagon_p^n|+1$, thereby concluding the proof.
\qed
\end{proof}

@@

\begin{lemma}[$\delta=3, \alpha =1$]\label{lem:d3a1}
	A deterministic unary oritatami system of $\delta=3$ and $\alpha = 1$ can fold into a terminal conformation of length at most $3n^2+4n+2$.
\end{lemma}
\begin{proof}
We will argue similarly to the $\delta=2$ case: when we stabilize beads outside $\hexagon_p^n$, we can only do so at points right next to one of the sides and even there only until we reach a corner. This yields the upper bound immediately.

### 2018-12-06 tt_intro.tex
Technical problems fixed

In this section, we consider Problem~\ref{prob:det_unary_length} at delay 1. 
Let $\Xi = (\Sigma, R, \delta, \alpha, \sigma, w)$ be a deterministic oritatami system of delay 1. 
For $i \geq 0$ let $C_i$ be the unique elongation of $\sigma$ by $w[1..i]$, that is, foldable by $\Xi$. Hence $C_0 = \sigma$.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.5\linewidth]{Fig/jordan_curve.png}
\caption{A tunnel divides the world into two. 
Getting through the tunnel leads us to another world. 
}
\label{fig:Jordan}
\end{figure}

@@

We now prove that in ``almost all'' tunnels is a troll domiciled and robs the transcript of binding capability (the original story is from \cite{Pratchett1992}). 
Originally, we tried to find a troll in every tunnel but failed; a troll seems to dislike the very first bead $w[1]$, or its property that only $\alpha$+1 beads around may take all hands of $w[1]$ thanks to the absence of its predecessor; any other bead must be surrounded by at least $\alpha+2$ beads in order to be free from free hand because a bead cannot bind with its predecessor or successor. 
We call a bead \textit{singular} if it is surrounded by only $\alpha+1$ beads but forms $\alpha$ bonds. 
Any bead but $w[1]$ can be singular because of their predecessor and successor. 
A tunnel is \textit{singular} if its entrance or exit is next to $w[1]$ that is singular. 
There can be at most 3 tunnels around one bead so that no more than 3 tunnels can be singular. 
A singular tunnel will be denoted with the superscript $\times$ like $t_s^\times$ or $t^\times$. 
In contrast, the notation without $\times$ such as $t_s$ and $t$ shall imply their non-singularity.

>>>>

In this section, we consider Problem~\ref{prob:det_unary_length} at delay 1 first and then at arity 1. 
Let $\Xi = (\Sigma, R, \delta, \alpha, \sigma, w)$ be a deterministic oritatami system of delay 1. 
For $i \geq 0$ let $C_i$ be the unique elongation of $\sigma$ by $w[1..i]$, that is, foldable by $\Xi$. Hence $C_0 = \sigma$.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.5\linewidth]{Fig/jordan_curve.png}
\caption{A tunnel divides the world into two. 
%Getting through the tunnel leads us to another world. 
}
\label{fig:Jordan}
\end{figure}

@@

We now prove that in ``almost all'' tunnels is a troll domiciled and robs the transcript of binding capability (the original story is from \cite{Pratchett1992}). 
Originally, we tried to find a troll in every tunnel but failed; a troll seems to dislike the very first bead $w[1]$, or its property that only $\alpha$+1 beads around may take all hands of $w[1]$ thanks to the absence of its predecessor; any other bead must be surrounded by at least $\alpha+2$ beads in order to be free from free hand because a bead cannot bind with its predecessor or successor. 
We call a bead \textit{singular} if it is surrounded by only $\alpha+1$ beads but forms $\alpha$ bonds. 
No bead but $w[1]$ can be singular because of their predecessor and successor. 
A tunnel is \textit{singular} if its entrance or exit is next to $w[1]$ that is singular. 
There can be at most 3 tunnels around one bead so that no more than 3 tunnels can be singular. 
A singular tunnel will be denoted with the superscript $\times$ like $t_s^\times$ or $t^\times$. 
In contrast, the notation without $\times$ such as $t_s$ and $t$ shall imply their non-singularity.

### 2018-12-06 tt_theorem.tex
Technical problems fixed

\begin{theorem}[$\delta = 1, \alpha = 4$]
Let $\Xi$ be a deterministic unary oritatami system of $\delta = 1, \alpha = 4$. It can yield only finite conformation of length $3n^2  + 3n + 1$.
\end{theorem}

\begin{proof}
Consider the moment when a bead $b$ is stabilized outside $\hexagon_O^n$ for the first time. 
See Fig.~\ref{TTT_a4_first}.
There is only one bead available to stabilize $b$ there, which is $b_1$. 
In order for $b$ to avoid nondeterminism, hence, none of the beads around should not attract $b$.

@@

@@

\begin{theorem}[$\delta = 1, \alpha = 3$]
Let $\Xi$ be a deterministic unary oritatami system of $\delta = 1, \alpha = 3$. It can yield only finite conformations of length $4n + 14$.
\end{theorem}

@@

      \draw[dotted, shift = (60:1)] (0:0) -- (-60:1);

@@

      \draw[dashed] (180:2) -- (0:2);
      \draw[dashed, shift=(60:1)] (180:2.5) -- (0:1.5);
       \draw[dashed, shift=(60:2)] (180:3) -- (0:1);
      \draw[->, blue] (60:0.1) -- (60:0.9);
      \draw[->, blue, shift=(60:1)] (0:0.1) -- (0:0.9);
       \draw[dotted, shift = (60:1)] (0:0) -- (-60:1);

@@

\begin{theorem}[$\delta= 1, \alpha=2$]
Let $\Xi$ be a deterministic unary oritatami system of $\delta = 1, \alpha = 2$.
It can yield an infinite structure, but the transcript folds into the zig-zag conformation after its $(27n^2 + 9n +1)$-th bead.
\end{theorem}

>>>>

\begin{theorem}[$\delta = 1, \alpha = 4$]
The terminal conformation of a deterministic unary oritatami system of $\delta = 1, \alpha = 4$ is of length at most $3n^2  + 3n + 1$.
\end{theorem}

\begin{proof}
Consider the moment when a bead, say $b$, is stabilized outside $\hexagon_O^n$ for the first time.
The bead must be bound a bead at the periphery of $\hexagon_O^n$ as depicted in Fig.~\ref{TTT_a4_first}.
%There is only one bead available to stabilize $b$ there, which is $b_1$. 
In order to avoid nondeterminism, the bead $b$ must not be attracted anyhow else by beads around.

@@

	\draw[dotted, thick] (1, 0) -- ++(120:1);

@@

\begin{theorem}[$\delta = 1, \alpha = 3$]
The terminal conformation of a deterministic unary oritatami system of $\delta = 1, \alpha = 3$ is of length at most $4n + 14$.
\end{theorem}

@@

      \draw[dotted, thick, shift = (60:1)] (0:0) -- (-60:1);

@@

      \draw[dashed] (180:2) -- (0:2);
      \draw[dashed, shift=(60:1)] (180:2.5) -- (0:1.5);
       \draw[dashed, shift=(60:2)] (180:3) -- (0:1);
      \draw[->, blue] (60:0.1) -- (60:0.9);
      \draw[->, blue, shift=(60:1)] (0:0.1) -- (0:0.9);
       \draw[dotted, thick, shift = (60:1)] (0:0) -- (-60:1);

@@

\begin{theorem}[$\delta= 1, \alpha=2$]
A deterministic unary oritatami system of $\delta = 1, \alpha = 2$ can fold into an infinite conformation, but its transcript folds into the zig-zag conformation (Fig.~\ref{TTT_zigzag}) after its $(27n^2 + 9n +1)$-th bead.
\end{theorem}
